An Introduction to Rocker: Docker Containers for R

### 2017-07-10 manuscript.Rmd
WIP

---
title: "An Introduction to Rocker: Docker Containers for R"
author:
  - name: Carl Boettiger
    affiliation: UC Berkeley 
    address:
    - ESPM Department, University of California,
    - 130 Mulford Hall Berkeley, CA 94720-3114, USA
    email:  cboettig@berkeley.edu
  - name: Dirk Eddelbuettel
    affiliation: Debian and R Projects
    address:
    - line 1
    - line 2
    email:  edd@debian.org
abstract: >
  We describe the Rocker project, which provides a widely used suite of Docker images
  with customized R environments for particular tasks. We discuss how this suite is
  organized and how these tools can increase portability, scaling, reproducibility,
  and convenience of R users and developers.
preamble: >
   \usepackage{longtable}
   \usepackage{svg}
output: rticles::rjournal_article
---

@@

## Using Rocker images

@@

## Use cases

There are several different but complementary reasons for creating the
containerized R environments provided by the Rocker project, including
convenience, community input, portability and scaling, reproducibility,
and instruction.  We describe these in turn before presenting the commands
required to deploy these environments in different use cases.

### Portability & cloud computing

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy an R envionment on a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers, or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

After running these two lines, the user may log into the server merely
by pasting its IP address or DNS name into a browser and entering their
password for user `rstudio`.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise (with many cloud providers, these two lines
of code can be run in a startup script without ever logging into the
server over ssh).  Some caveats and additional configuration options are
discussed in the example section below.  The user can select a Rocker
container which contains the required software pre-installed, such as
`rocker/tidyverse`, which provides the popular `tidyverse` suite of R
packages \citep{tidyverse}, or even use their own containers which can
extend the Rocker containers, as discussed below.  This workflow saves
considerable time relative to installing each package from scratch,
which can take over an hour to reconstruct the environment provided by
some rocker containers.  As many commericial cloud providers charge by
the hour, a fast and simple mechanism to deploy the desired environment
in the cloud is crucial to making this approach practical for more
exploratory analyses.

<!--
FIXME describe other modes of running containers, e.g. R terminal

Another illustration of this portability is with regard to Continuous
Integration servers such as Travis <https://travis.com> or Circle-CI
<https://circle-ci.com>.  While support for the R language on Travis has
greatly improved recently (\citep{r-travis}), very customized environments
can still be difficult to run (particularly for testing scripts not laid
out in the R package structure).

>>>>

---
title: "An Introduction to Rocker: Docker Containers for R"
author:
  - name: Carl Boettiger
    affiliation: UC Berkeley 
    address:
    - ESPM Department, University of California,
    - 130 Mulford Hall Berkeley, CA 94720-3114, USA
    email:  cboettig@berkeley.edu
  - name: Dirk Eddelbuettel
    affiliation: Debian and R Projects
    address:
    - line 1
    - line 2
    email:  edd@debian.org
abstract: >
  We describe the Rocker project, which provides a widely used suite of Docker images
  with customized R environments for particular tasks. We discuss how this suite is
  organized and how these tools can increase portability, scaling, reproducibility,
  and convenience of R users and developers.
preamble: >
   \usepackage{longtable}
output: rticles::rjournal_article
---

<!-- Notes to Dirk: 
     Some of this material may be more appropriate for documentation elsewhere, appearing here in only abbreviated form.  
     To get started, I've just tried to err on the side of comprehensive, particularly since our documentation in places is sparse
     or dated.  Feel free to flag any areas that should be moved in part or whole to documentation.  -->

@@

## Design principles & use cases

- Portable: from laptop to cloud
- Sandboxed: experiment without touching local system
- Optimized: community-tweaked configurations
- Interfaces: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
- Extensible: interactively and persistent community extensions
- Versioned:

research use, instructional use

### Portability: From laptop to cloud

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R envionment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers, or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

After running these two lines, the user may log into the server merely
by pasting its IP address or DNS name into a browser and entering the
appropriate password.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise.

The user can select a Rocker
container which contains the required software pre-installed, such as
`rocker/tidyverse`, which provides the popular `tidyverse` suite of R
packages \citep{tidyverse}, or even use their own containers which can
extend the Rocker containers, as discussed below.  This workflow saves
considerable time relative to installing each package from scratch,
which can take over an hour to reconstruct the environment provided by
some rocker containers.  As many commericial cloud providers charge by
the hour, a fast and simple mechanism to deploy the desired environment
in the cloud is crucial to making this approach practical for more
exploratory analyses.

### Interfaces

@@
